Chemistry of the Orphan Stream I: Identifying Stream Members \\ from Low-Resolution Spectroscopy

### 2012-09-16 2012-oss-mrs/draft-oss-mrs.tex
Awaiting GDC velocity dispersions

\title{Chemistry of the Orphan Stream I: Identifying Stream Members from Low-Resolution Spectroscopy}

@@

\begin{abstract}
We present coordinates and properties of candidate K giant members in the Orphan Stream, which have been identified from low resolution spectroscopic data taken with AAOmega on the Anglo-Astronomical Telescope. With modest S/N and independent cuts in photometry, kinematics, gravity and metallicity we yield self-consistent, highly probable stream members. Our data indicates an overall stream metallicity of [Fe/H] $= -1.71$ with an intrinsic dispersion of $\sigma = 0.56$\,dex which is more metal-rich than previously found and unbiased by spectral type. The metallicity dispersion in our stream sample is indicative that the Orphan Stream has a dSph origin. We find a revised stream distance of $23.5 \pm 1.0$\,kpc near the celestial equator, and our kinematic signature peaks at $V_{GSR} = 00.0$\,km s$^{-1}$. The observed velocity dispersion of our highest probability members is consistent with arising from the velocity uncertainties alone, indicating that, at least along the line-of-sight, the Orphan Stream is kinematically quite cold. We highlight likely members for high resolution spectroscopic follow-up.
\end{abstract}

@@

The Orphan Stream was independently detected by both \citet{Grillmair;Dionatos_2006} and \citet{Belokurov;et-al_2006}, and is distinct from other substructures in the halo. The stream stretches over $60\,^\circ$ in the sky, has a low surface brightness and a narrow stream width of only $\sim2\,^\circ$ and as the name suggests, the parent object largely remains a mystery. The stream extends past the celestial equator \--- outside the SDSS footprint \--- but has not been detected in existing southern surveys \citep{Newberg;et-al_2010}. Whilst the parent system remains elusive, significant effort has been placed on associating the stream with known Milky Way satellites \citep{Zucker;et-al_2006, Fellhaur;et-al_2007,Jin;Lynden_Bell_2007,Sales;et-al_2008}. In contrast, there has been relatively limited observational work on the Orphan Stream itself other than the original discovery papers \citep{Grillmair;Dionatos_2006, Belokurov;et-al_2006, Belokurov;et-al_2007} and the work of \citet{Newberg;et-al_2010}. This is largely to be expected given the absence of multi-band photometry in the southern sky and the low total luminosity of the stream, making it difficult to reliably seperate Orphan Stream members from halo stars. Further understanding the extent of the Orphan Stream awaits the SkyMapper and Pan-STARRS photometric surveys \citep{Keller;et-al_2007, Hodapp;et-al_2004}.

As \citet{Sales;et-al_2008} points out, there is a natural observational bias towards more massive and recent mergers like Sagittarius. Consequently, the fainter end of this substructure distribution has yet to be fully recovered, or thoroughly examined. Interestingly, there are indications that  some fainter substructures like the Orphan Stream and the Palomar 5 tidal tails \citep{Odenkirchen;et-al_2009} have orbits which seem to be best-fit by Milky Way models with nearly 60\% less mass \citep{Newberg;et-al_2010} than generally reported by \citet{Xue;et-al_2008} and \citet{Koposov;et-al_2010}. Such a discrepancy in the mass of the Milky Way is troublesome. More complete photometric and kinematic maps of these low total luminosity streams may provide the best test as to whether this mass discrepancy is real, or an artefact of incomplete observations. Whilst the full spatial extent of the stream remains incomplete, we can examine the detailed chemistry of its members and investigate the stream history, as well as make predictions about the nature of the host system.

In this letter we present a detailed, self-consistent analysis to identify K giant members of the Orphan Stream. Using our selection method we have catalogued the locations of ten highly probable Orphan Stream candidates, all worthy of high-resolution spectroscopic follow up. In the following section we outline our photometric target selection. In \S\ref{sec:observations} we describe the low-resolution spectroscopic observations. The data analysis, including stream identification, is discussed in \S\ref{sec:analysis} and in \S\ref{sec:conclusions} the conclusions, predictions and future work are presented.
\newline

@@

We have targeted K giant members of the Orphan Stream in order to investigate their detailed chemistry. Because K-giants are difficult to unambiguously detect from photometry alone, low-resolution spectroscopy is required to estimate stellar parameters and determine radial velocities. The Orphan Stream has a low spatial over-density on the order of $\sim10\%$, which makes it difficult to separate stream members from halo stars. However, there is a well described distance gradient along the stream \citep{Belokurov;et-al_2007, Newberg;et-al_2010} which provides an indication on where we should focus our spectroscopic efforts.

@@

\subsection{Kinematics}
Radial velocities were measured by cross-correlating our normalised spectra against a K-giant synthetic template with a temperature of 4500\,K, $\log{g}$ = 1 and $[\mbox{Fe/H}] = -1.5$ across the range $845 \leq \lambda \leq 870$\,nm. Heliocentric velocities were translated to the galactic rest frame by adopting the local standard of rest velocity as 220\,km s$^{-1}$ towards $(l, b) = (53\,^\circ, 25\,^\circ)$ \citep{Kerr;Lynden-Bell_1986, Mihalas;Binney_1981}\footnote{Where $V_{GSR} = V_{\textsc{helio}} + 220\sin{l}\cos{b} + 16.5\times[\sin{b}\sin{25} + \cos{b}\cos{25}\cos{(l - 53)}]$}. \citet{Newberg;et-al_2010} presented a comprehensive kinematic dataset on the Orphan Stream using BHB stars and employed a slightly different galactocentric transformation. This systematic difference equates to our galactocentric rest frame velocities being on average $\sim5$\,km s$^{-1}$ lower than those of \citet{Newberg;et-al_2010}.

@@

In this region \citet{Newberg;et-al_2010} found the Orphan Stream to have $V_{GSR} = 110$ km s$^{-1}$ from BHB stars, approximately $105$\,km s$^{-1}$ on our scale. To isolate potential Orphan Stream members we have nominated a relatively wide kinematic selection region between $75 \leq V_{GSR} \leq 135$ km s$^{-1}$, which yields 25 Orphan Stream candidates. The typical uncertainty in our kinematics is $\pm{}3.5$\,km s$^{-1}$, although for our fainter targets this quickly increases above 10\,km s$^{-1}$.

@@

We have measured the equivalent widths of the gravity-sensitive Mg\,\textsc{I} line at 8807 \AA{} to distinguish dwarfs from giants \citep{Battaglia;Starkenburg_2012}. At a given temperature (or $g \-- r$) and metallicity, giant stars present narrower Mg\,\textsc{I} absorption lines than their dwarf contaminants. Given our target selection, our sample is likely to contain many more dwarfs than giants. In some cases no Mg\,\textsc{I} 8807 \AA{} line was apparent, so an upper limit was estimated based on the S/N. In these cases the candidate was considered a ``non-dwarf" because we cannot exclusively rule out a metal-poor sub-giant with this criterion alone. For these purposes we are only looking for a simple indication as to whether a star is likely a dwarf or not. Figure \ref{fig:ew-mg} illustrates the trend with EW$_{\lambda8807}$ against SDSS de-reddened\footnote{All magnitudes presented in this letter are de-reddened using the \citet{Schlegel;Finkbeiner;Davis_1998} dust maps.} $g \-- r$, illustrating the dominant upper dwarf branch we wish to exclude. The adopted dwarf/giant separation line is shown. Admittedly this selection criterion is not particularly strict, but we are allowing multiple inclusive cuts and taking the intersection of these criteria to identify Orphan Stream giants.

\begin{figure}[h]
	\includegraphics[width=\columnwidth]{./figures/ew-mg.pdf}
	\caption{De-reddened $g \-- r$ against the measured equivalent width of the Mg\,\textsc{I} transition at 8807 \AA{}. Dwarf contaminants occupy the more populous upper branch. Our separation line between dwarfs and giants is shown in green.}
	\label{fig:ew-mg}
\end{figure}

@@

Our dwarf/giant separation line in Figure \ref{fig:ew-mg} yields 425 potential giants. This fraction of giants ($\sim{}44\%$) is much larger than what would be expected given our selection criterion (e.g. see \citeauthor{Casey;et-al_2012} \citeyear{Casey;et-al_2012}a). Hence, our gravity criterion is extremely inclusive and susceptible to dwarf contamination. Upon taking the intersection of our kinematic and gravity selections, we deduce eighteen stars that appear likely Orphan Stream K giants.

@@

We have measured the equivalent widths of the Ca\,\textsc{II} near-infrared triplet lines for stars that meet our kinematic and surface gravity criteria. After correcting for luminosity, the equivalent width of the Ca\,\textsc{II} triplet lines provide a good indication of the overall metallicity of a RGB star \citep{Amandroff;Da_Costa_1991}. We have employed the \citet{Starkenburg;et-al_2010} relationship and corrected for luminosity in $g$ against the horizontal branch magnitude at $g_{HB}$ = 17.1 \citep{Newberg;et-al_2010}. Strictly speaking, the Ca\,\textsc{II}\---[Fe/H] calibration is only valid for stars brighter than the horizontal branch, although the relationship only becomes significantly inappropriate near $g \-- g_{HB} \sim +1$ \citep{Saviane;et-al_2012}. Many of our candidates are fainter than this valid luminosity range, and therefore they should not be excluded solely because of their derived metallicities. Stars fainted than $g_{HB}$ will have slightly lower metallicities than predicted by our Ca\,\textsc{II}\---[Fe/H] relationship, and for these stars we will only use metallicities to assign a relative qualitative likelihood for stream membership.
Given a distance to the Orphan Stream, we can also estimate a star's metallicity through isochrone fitting. We have used a 10 Gyr \citet{Girardi;et-al_2008} isochrone at 21.4\,kpc \citep{Newberg;et-al_2010} and found metallicities for all eighteen likely stream members from their best-fitting isochrone (see Figure \ref{fig:feh}). Derived metallicities from Ca\,\textsc{II} line strengths and isochrone fitting that are consistent (within $\pm0.3$\,dex) indicates these measurements are reliable, and that these stars are indeed at a distance of $\sim21.4$\,kpc. We find ten highly likely stream members with consistently derived metallicities. They fall within the shaded region illustrated in Figure \ref{fig:feh}.

@@

\begin{figure}[h]
	\includegraphics[width=\columnwidth]{./figures/cmd.pdf}
	\caption{Color magnitude diagram showing our observed candidates (grey). Observations fulfilling kinematic and gravity cuts are colored by their metallicity, and those with upper limits for surface gravity are marked as triangles ($\triangledown$). Highly probable stream members with consistent line strength- and isochrone-derived metallicities are circled. Relevant 10 Gyr \citet{Girardi;et-al_2008} isochrones at $[\mbox{Fe/H}] = -1.5$ (dotted), $-2.0$ (dashed) at 21.4\,kpc \citep{Newberg;et-al_2010} are shown, as well as our best-fitting 10 Gyr isochrone of $[\mbox{Fe/H}] = -1.71$ at 23.5\,kpc (solid).}
	\label{fig:cmd}
\end{figure}

@@

We have presented a detailed analysis to isolate individual Orphan Stream K giants from low-resolution spectroscopy using a combination of photometric, kinematic, gravity, and metallicity information. Although each individual criterion is likely to induce some level contamination, their intersection reveals 10 highly probable, self-consistent, Orphan Stream K giants.  We deduce a median stream metallicity of $[\mbox{Fe/H}] = -1.71$ and find an intrinsically wide metallicity spread of $\sigma = 0.56$\,dex, indicative of a dSph origin. Unlike other stellar tracers, K-type giants can exist at all metallicities, hence our derived metallicity spread is likely representative of the true stream metallicity distribution function. Recall that the metallicity determination was performed after kinematic and gravity cuts, and our most probable members lay perfectly on a 10\,Gyr isochrone of $[\mbox{Fe/H}] = -1.71$. Our stream members indicate a revised distance to the stream of $23.5 \pm 1.0$\,kpc at $(l, b) = (250\,^\circ$, $50\,^\circ)$, slightly farther than that deduced by \citet{Newberg;et-al_2010}.

@@

The K giants presented here can provide great insight into the chemistry, and history of the Orphan Stream. High-resolution spectroscopic observations have been taken for our most highly probable members and a detailed chemical analysis will be presented in a forthcoming paper (Casey et al. 2012c, in prep.). Detailed chemical abundances can help determine both the nature of the host system, and allow us to compare peculiar chemical signatures with those of the known Milky Way satellites in order to associate likely parents. However, at least for the moment, the Orphan Stream remains appropriately named.

>>>>

\title{Chemistry of the Orphan Stream I: Identifying Stream Members \\ from Low-Resolution Spectroscopy}

@@

\begin{abstract}
We present coordinates and properties of candidate K giant members in the Orphan Stream, which have been identified from low resolution spectroscopic data taken with AAOmega on the Anglo-Astronomical Telescope. With modest S/N and independent cuts in photometry, kinematics, gravity and metallicity we yield self-consistent, highly probable stream members. Our data indicates an overall stream metallicity of [Fe/H] $= -1.71$ with an intrinsic dispersion of $\sigma = 0.56$\,dex which is more metal-rich than previously found and unbiased by spectral type. The metallicity dispersion in our stream sample suggests that the unidentified Orphan Stream parent is a dSph satellite. We find a revised stream distance of $23.5 \pm 1.0$\,kpc near the celestial equator, and our kinematic signature peaks at $V_{GSR} = 00.0$\,km s$^{-1}$. The observed velocity dispersion of our highest probability members is consistent with arising from the velocity uncertainties alone, indicating that at least along the line-of-sight, the Orphan Stream is kinematically quite cold. We highlight likely members for high resolution spectroscopic follow-up.
\end{abstract}

@@

The Orphan Stream was independently detected by both \citet{Grillmair;Dionatos_2006} and \citet{Belokurov;et-al_2006}, and is distinct from other substructures in the halo. The stream stretches over $60\,^\circ$ in the sky, has a low surface brightness and a narrow stream width of only $\sim$2\,$^\circ$ and as the name suggests, the parent object largely remains a mystery. The stream extends past the celestial equator \--- outside the SDSS footprint \--- but has not been detected in existing southern surveys \citep{Newberg;et-al_2010}. Whilst the parent system remains elusive, significant effort has been placed on associating the stream with known Milky Way satellites \citep{Zucker;et-al_2006, Fellhaur;et-al_2007,Jin;Lynden_Bell_2007,Sales;et-al_2008}. In contrast, there has been relatively limited observational work on the Orphan Stream itself other than the original discovery papers \citep{Grillmair;Dionatos_2006, Belokurov;et-al_2006, Belokurov;et-al_2007} and the work of \citet{Newberg;et-al_2010}. This is largely to be expected given the absence of multi-band photometry in the southern sky and the low total luminosity of the stream, making it difficult to reliably separate Orphan Stream members from halo stars. Further understanding on the extent of the stream awaits the SkyMapper and Pan-STARRS photometric surveys \citep{Keller;et-al_2007, Hodapp;et-al_2004}.

As \citet{Sales;et-al_2008} point out, there is a natural observational bias towards more massive and recent mergers like Sagittarius. Consequently, the fainter end of this substructure distribution has yet to be fully recovered, or thoroughly examined. Interestingly, there are indications that  some fainter substructures like the Orphan Stream and the Palomar 5 tidal tails \citep{Odenkirchen;et-al_2009} have orbits which seem to be best-fit by Milky Way models with nearly 60\% less mass \citep{Newberg;et-al_2010} than generally reported by \citet{Xue;et-al_2008} and \citet{Koposov;et-al_2010}. Such a discrepancy in the mass of the Milky Way is troublesome. More complete photometric and kinematic maps of these low total luminosity streams may provide the best test as to whether this mass discrepancy is real, or an artefact of incomplete observations. Whilst the full spatial extent of the stream remains incomplete, we can examine the detailed chemistry of its members and investigate the stream history, as well as make predictions about the nature of the host system.

In this letter we present a detailed, self-consistent analysis to identify K giant members of the Orphan Stream. Using our selection method we have catalogued the locations of ten highly probable Orphan Stream candidates, all worthy of high-resolution spectroscopic follow up. In the following section we outline our photometric target selection. In \S\ref{sec:observations} we describe the low-resolution spectroscopic observations. The data analysis, including stream identification, is discussed in \S\ref{sec:analysis} and in \S\ref{sec:conclusions} the conclusions, predictions and future work are presented.

@@

We have targeted K giant members of the Orphan Stream in order to investigate their detailed chemistry. Because K-giants are difficult to unambiguously detect from photometry alone, low-resolution spectroscopy is required to estimate stellar parameters and determine radial velocities. The Orphan Stream has a low spatial over-density on the order of $\sim$10\%, which makes it difficult to separate stream members from halo stars. However, there is a well described distance gradient along the stream \citep{Belokurov;et-al_2007, Newberg;et-al_2010} which provides an indication on where we should focus our spectroscopic efforts.

@@

\subsection{Kinematics}
Radial velocities were measured by cross-correlating our normalised spectra against a K-giant synthetic template with a temperature of 4500\,K, $\log{g}$ = 1 and $[\mbox{Fe/H}] = -1.5$ across the range $845 \leq \lambda \leq 870$\,nm. Heliocentric velocities were translated to the galactic rest frame by adopting the local standard of rest velocity as 220\,km s$^{-1}$ towards $(l, b) = (53\,^\circ, 25\,^\circ)$ \citep{Kerr;Lynden-Bell_1986, Mihalas;Binney_1981}\footnote{Where $V_{GSR} = V_{\textsc{helio}} + 220\sin{l}\cos{b} + 16.5\times[\sin{b}\sin{25} + \cos{b}\cos{25}\cos{(l - 53)}]$}. \citet{Newberg;et-al_2010} presented a comprehensive kinematic dataset on the Orphan Stream using BHB stars and employed a slightly different galactocentric transformation. This systematic difference equates to our galactocentric rest frame velocities being on average $\sim$5\,km s$^{-1}$ lower than those of \citet{Newberg;et-al_2010}.

@@

In this region \citet{Newberg;et-al_2010} found the Orphan Stream to have $V_{GSR} = 110$ km s$^{-1}$ from BHB stars, approximately $105$\,km s$^{-1}$ on our scale. To isolate potential Orphan Stream members we have nominated a relatively wide kinematic selection region between $75 \leq V_{GSR} \leq 135$ km s$^{-1}$, which yields twenty-five Orphan Stream candidates. The typical uncertainty in our kinematics is $\pm{}5.0$\,km s$^{-1}$, although for our fainter targets this quickly increases above 10\,km s$^{-1}$.

@@

We have measured the equivalent widths of the gravity-sensitive Mg\,\textsc{I} line at 8807 \AA{} to distinguish dwarfs from giants \citep{Battaglia;Starkenburg_2012}. At a given temperature (or $g - r$) and metallicity, giant stars present narrower Mg\,\textsc{I} absorption lines than their dwarf contaminants. Given our target selection, our sample is likely to contain many more dwarfs than giants. In some cases no Mg\,\textsc{I} 8807 \AA{} line was apparent, so an upper limit was estimated based on the S/N. In these cases the candidate was considered a ``non-dwarf" because we cannot exclusively rule out a metal-poor sub-giant with this criterion alone. For these purposes we are only looking for a simple indication as to whether a star is likely a dwarf or not. Figure \ref{fig:ew-mg} illustrates the trend with EW$_{\lambda8807}$ against SDSS de-reddened\footnote{All magnitudes presented in this letter are de-reddened using the \citet{Schlegel;Finkbeiner;Davis_1998} dust maps.} $g - r$, illustrating the dominant upper dwarf branch we wish to exclude. The adopted dwarf/giant separation line is shown. Admittedly this selection criterion is not particularly strict, but we are allowing multiple inclusive cuts and taking the intersection of these criteria to identify Orphan Stream giants.

\begin{figure}[h]
	\includegraphics[width=\columnwidth]{./figures/ew-mg.pdf}
	\caption{SDSS $g - r$ against the measured equivalent width of the Mg\,\textsc{I} transition at 8807\,\AA{}. Dwarf contaminants occupy the more populous upper branch. Our separation line between dwarfs and giants is shown in green.}
	\label{fig:ew-mg}
\end{figure}

@@

Our dwarf/giant separation line in Figure \ref{fig:ew-mg} yields 425 potential giants. This fraction of giants ($\sim$44\%) is much larger than what would be expected given our selection criterion (e.g. see \citeauthor{Casey;et-al_2012} \citeyear{Casey;et-al_2012}a). Hence, our gravity criterion is extremely inclusive and susceptible to dwarf contamination. Upon taking the intersection of our kinematic and gravity selections, we deduce eighteen stars that appear likely Orphan Stream K giants.

@@

We have measured the equivalent widths of the Ca\,\textsc{II} near-infrared triplet lines for stars that meet our kinematic and surface gravity criteria. After correcting for luminosity, the equivalent width of the Ca\,\textsc{II} triplet lines provide a good indication of the overall metallicity of a RGB star \citep{Amandroff;Da_Costa_1991}. We have employed the \citet{Starkenburg;et-al_2010} relationship and corrected for luminosity in $g$ against the horizontal branch magnitude at $g_{HB}$ = 17.1 \citep{Newberg;et-al_2010}. Strictly speaking, the Ca\,\textsc{II}\---[Fe/H] calibration is only valid for stars brighter than the horizontal branch, although the relationship only becomes significantly inappropriate near $g \-- g_{HB} \sim +1$ \citep{Saviane;et-al_2012}. Many of our candidates are fainter than this valid luminosity range, and therefore they should not be excluded solely because of their derived metallicities. Stars fainted than $g_{HB}$ will have slightly lower metallicities than predicted by our Ca\,\textsc{II}\---[Fe/H] relationship, and for these stars we will only use metallicities to assign a relative qualitative likelihood for stream membership.
Given a distance to the Orphan Stream, we can also estimate a star's metallicity through isochrone fitting. We have used a 10\,Gyr \citet{Girardi;et-al_2008} isochrone at 21.4\,kpc \citep{Newberg;et-al_2010} and found metallicities for all eighteen likely stream members from their best-fitting isochrone (see Figure \ref{fig:feh}). Derived metallicities from Ca\,\textsc{II} line strengths and isochrone fitting that are consistent (within $\pm0.3$\,dex) indicates these measurements are reliable, and that these stars are indeed at a distance of $\sim$21.4\,kpc. We find ten highly likely stream members with consistently derived metallicities. They fall within the shaded region illustrated in Figure \ref{fig:feh}.

@@

\begin{figure}[h]
	\includegraphics[width=\columnwidth]{./figures/cmd.pdf}
	\caption{Color magnitude diagram showing our observed candidates (grey). Observations fulfilling kinematic and gravity cuts are colored by their metallicity, and those with upper limits for surface gravity are marked as triangles ($\triangledown$). Highly probable stream members with consistent line strength- and isochrone-derived metallicities are circled. Relevant 10\,Gyr \citet{Girardi;et-al_2008} isochrones at $[\mbox{Fe/H}] = -1.5$ (dotted), $-2.0$ (dashed) at 21.4\,kpc \citep{Newberg;et-al_2010} are shown, as well as our best-fitting 10\,Gyr isochrone of $[\mbox{Fe/H}] = -1.71$ at 23.5\,kpc (solid).}
	\label{fig:cmd}
\end{figure}

@@

We have presented a detailed analysis to isolate individual Orphan Stream K giants from low-resolution spectroscopy using a combination of photometric, kinematic, gravity, and metallicity information. Although each individual criterion is likely to induce some level contamination, their intersection reveals ten highly probable, self-consistent, Orphan Stream K giants.  We deduce a median stream metallicity of $[\mbox{Fe/H}] = -1.71$ and find an intrinsically wide metallicity spread of $\sigma = 0.56$\,dex, indicative of a dSph origin. Unlike other stellar tracers, K-type giants can exist at all metallicities, hence our derived metallicity spread is likely representative of the true stream metallicity distribution function. Recall that the metallicity determination was performed after kinematic and gravity cuts, and our most probable members lay perfectly on a 10\,Gyr isochrone of $[\mbox{Fe/H}] = -1.71$. Our stream members indicate a revised distance to the stream of $23.5 \pm 1.0$\,kpc at $(l, b) = (250\,^\circ$, $50\,^\circ)$, slightly farther than that deduced by \citet{Newberg;et-al_2010}.

@@

The K giants presented here can provide great insight into the chemistry, and history of the Orphan Stream. High-resolution spectroscopic observations have been taken for our most highly probable members and a detailed chemical analysis will be presented in a forthcoming paper (Casey et al. 2012c, in prep.). Detailed chemical abundances can help determine both the nature of the host system before it's discovered, and allows us to compare peculiar chemical signatures with those of the known Milky Way satellites in order to associate likely parents. However, at least for the moment, the Orphan Stream remains appropriately named.
